---
title: "cBioPortal API Workshop"
author: "Xiaoni"
date: "6/5/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Introduction
This vignette lays out the two main user-facing functions for downloading and representing data from the cBioPortal API. cBioDataPack makes use of the legacy distribution data method in cBioPortal (via tarballs). cBioPortalData allows for a more flexibile approach to obtaining data based on several available parameters including available molecular profiles.

##Installation

```{r}
#BiocManager::install('cBioPortalData')
library(cBioPortalData)
library(AnVIL)
```

##Initialize the API Client
###cBioPortalData: Obtain data from the cBioPortal API


```{r}
cbio <- cBioPortal()
cbio
```

###Compute some simple stats
1.how many studies are there?
2.how many cancer types do they span?
3.how many samples in total?
4.which study has the largest number of samples?


###Pull data from the API endpoint using cBioPortalData

```{r}
resp <- cbio$getAllStudiesUsingGET()
resp
```

####parseResponse
```{r}
parsedResponse <- httr::content(resp)
cat("Number of elements in the response:", length(parsedResponse))

```
Since we know in this case that each element represents a study, we can answer question 1:

```{r question 1}
cat("Answer 1: There are", length(parsedResponse), "studies in cBioPortalData")
```


```{r}
studies <- getStudies(cbio)
studies
```


```{r question 1 alternate}
dim(studies)
cat("Answer 1: There are", nrow(studies), "studies in cBioPortal")
colnames(studies)
```

```{r question 2}
cat("Answer 2: The studies span", length(unique(studies$cancerTypeId)), "cancer types")
```

```{r question 3}
cat("Answer 3: There are", sum(studies$allSampleCount), "samples in cBioPortal")

```

```{r question 4}
cat("Answer 4: The study with the most samples is", studies[which.max(studies$allSampleCount), "name"][[1]])

```



####List all function in a package
```{r}
ls("package:cBioPortalData")
```

```{r}
browseVignettes(package = 'cBioPortalData')
```



###Visualize data

####Transform data using tidyverse
Get the top 20 counts per cancer type
```{r}
cancerTypeCounts_T <-                                                  # assign results
  studies %>%                                                          # %>% is the pipe operator
  group_by(cancerTypeId) %>%                                           # and group by cancer type
  summarise(totalSamples=sum(allSampleCount), .groups = 'drop') %>%    # sum allSampleCount, add column
  arrange(desc(totalSamples)) %>%                                      # sort by totalSamples
  top_n(20)                                                            # take top 20

cancerTypeCounts_T 
```


```{r filter mixed}
cancerTypeCounts <- 
  studies %>%
  filter(cancerTypeId != 'mixed') %>%
  group_by(cancerTypeId) %>%
  summarise(totalSamples=sum(allSampleCount), .groups = 'drop') %>%
  arrange(desc(totalSamples)) %>%
  top_n(20)

cancerTypeCounts 
```


```{r}
library(ggplot2)
cancerTypeCounts <- cancerTypeCounts %>% arrange(cancerTypeCounts$totalSamples)
par(mar = c(4,6,1,1))
barplot(cancerTypeCounts$totalSamples,
        names = cancerTypeCounts$cancerTypeId,
        main = "Number of samples by top 20 Primary Sites",
        horiz = TRUE,
        las = 1)

```


###Example 1: Get data from Prostate Adenocarcinoma (TCGA, Cell 2015)
```{r}
## load packages:
library(cBioPortalData)
library(httr)
library(dplyr)
library(stringr)
library(biomaRt)
library(ggplot2)
studies = as.data.frame(getStudies(cBioPortal()))
head(studies)
```
#### list all prostate cancer studies available

```{r}
studies[grep('Prostate.*', studies$name), c('name', 'studyId')] 

##go for Prostate Adenocarcinoma (TCGA, Cell 2015); prad_tcga_pub
## list all study (ID) participants
all.patients = cbio$getAllPatientsInStudyUsingGET(studyId = 'prad_tcga_pub')
all.patients = httr::content(all.patients, as = 'parsed')

## convert to data frame; easier downstream handling
patients_dataframe = data.frame(matrix(
  unlist(all.patients),
  nrow = length(all.patients),
  byrow = T
))
colnames(patients_dataframe) = c('Identifier', 'Patient_ID', 'Study_ID')
head(patients_dataframe)


all_samples = allSamples(cbio, studyId = 'prad_tcga_pub')
head(all_samples)
```

###look into clinical attributes (data)
```{r}
patient.x.clinics = cbio$getAllClinicalDataOfPatientInStudyUsingGET(studyId = 'prad_tcga_pub',
                                                                    patientId = 'TCGA-VP-A87C')
patient.x.clinics = httr::content(patient.x.clinics)
patient.x.clinics = data.frame(matrix(
  unlist(patient.x.clinics),
  nrow = length(patient.x.clinics),
  byrow = T))

## delete first (unesseccary) column
patient.x.clinics$X1 = NULL

colnames(patient.x.clinics) = c('Patient_ID', 'Study_ID', 'Attribute', 'Value')
head(patient.x.clinics)

```

```{r}
patient.x.sample = cbio$getAllClinicalDataOfSampleInStudyUsingGET(studyId = 'prad_tcga_pub',
sampleId = 'TCGA-VP-A87C-01') ## compare to above here
patient.x.sample = httr::content(patient.x.sample)
patient.x.sample = data.frame(matrix(
  unlist(patient.x.sample),
  nrow = length(patient.x.sample),
  byrow = T)) 
patient.x.sample$X1 = NULL
patient.x.sample$X2 = NULL
patient.x.sample$X3 = NULL
colnames(patient.x.sample) = c('Patient_ID', 'Study_ID', 'Attribute', 'Value')
head(patient.x.sample) 
```

#### retrieve all the clinical data from all study participants
```{r}
## get all the clinical data for all patients
all.clinics = cbio$getAllClinicalDataInStudyUsingGET(studyId = 'prad_tcga_pub')
all.clinics = httr::content(all.clinics)
all.clinics = data.frame(matrix(
  unlist(all.clinics),
  nrow = length(all.clinics),
  byrow = T)) ## convert again, easier handling
all.clinics[, c(1, 2, 3)] = NULL
head(all.clinics)

all_clinical_data = clinicalData(cbio, studyId = 'prad_tcga_pub')
head(all_clinical_data)
```

####plot Age at diagnosis

```{r}
ggplot() +
  geom_bar(
    data = all_clinical_data,
    aes(x = as.numeric(AGE)),
    color = 'royalblue',
    fill = 'royalblue',
    na.rm = T,
    width = 0.8) +
  theme_bw() +
  scale_x_continuous(breaks = seq(44, 76, 2)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 50)) +
  theme(axis.text.x = element_text(angle = -45)) +
  labs(x = '', y = '', title = 'Diagnosis Age')

```
##### race category

```{r}
race = as.data.frame(table(all_clinical_data$RACE))
ggplot(race, aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) + theme_void()
```
#### look at number of mutations vs fraction genome altered:
```{r}
FGA = all.clinics[all.clinics$X6 == 'FRACTION_GENOME_ALTERED', ] ## subset dataframe
mutation.count = all.clinics[all.clinics$X6 == 'MUTATION_COUNT', ] ## subset dataframe
data.figure1 = merge(FGA[,c(1, 4)],
                     mutation.count[,c(1,4)],
                     by.x = 'X4',
                     by.y = 'X4',
                     all = T) ## merge the two data frame to prepare for plot
colnames(data.figure1) = c('patient', 'FGA', 'mutation.count')


ggplot(data.figure1, aes(x = as.numeric(FGA),
                         y = as.numeric(mutation.count))) +
  geom_jitter() +
  theme_bw() +
  theme(legend.position = 'none', aspect.ratio = 1) +
  scale_y_continuous(limits = c(0, 750), expand = c(0,0), breaks = seq(0, 800, by = 100)) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), breaks = seq(0, 1, by = 0.1)) +
  labs(x = 'Fraction Genome Altered', y = '# of mutations', title = 'Mutation count VS FGA')
```

###Turning into molecular data
```{r}
genomic_parameters = molecularProfiles(cbio,
                                       studyId = 'prad_tcga_pub',
                                       projection = c('SUMMARY'))

print(genomic_parameters[, 'molecularProfileId'])
print(genomic_parameters[, 'molecularAlterationType'])
```
#### look into one specific patient and retrieve the mutational spectrum
```{r}
mart.object <- useMart("ensembl", 
                       dataset = "hsapiens_gene_ensembl")
genes_ids <- getBM(mart = mart.object,
                   attributes = c("hgnc_symbol", "entrezgene_id"))

genes_ids = unique(as.vector(genes_ids$entrezgene_id), na.rm = T)
all.patient.mutations = molecularData(
  api = cbio,
  molecularProfileId = 'prad_tcga_pub_mutations',
  entrezGeneIds = genes_ids,
  sampleIds = paste0(patients_dataframe$Patient_ID, '-01'),
  check = T
)
head(all.patient.mutations)


## look into one specific patient:
all.patient.mutations[all.patient.mutations$patientId == 'TCGA-VP-A87C', ]
```
##What is MultiAssayExperiment?

```{r}
browseVignettes(package = "MultiAssayExperiment")
```

#Example 2:MAE

####Brief explanation what an MAE is:
The essence of a MultiAssayExperiment: * The essence of MultiAssayExperiment is the following: Imagine
you have a bookshelf;
* The whole bookshelf is one particular study; then image the levels within the bookshelf * every levels in
this bookshelf is one particular assay or experiment (e.g. CopyNumber Calls)
* Those Assays/Experiments are stored as lists within the MultiAssayExperiment Object. you can combine
different experiments on one sample/study and commonly work with them
experiments may be measures of mRNA, copyNumber Alterations, and mutations calls (among others)


####Example:
On Lung Invasisve Adenocarcinomas (LUAD) from TCGA by creating a MAE object with luad_tcga data from cBIO

#####load libraries
```{r}
library(cBioPortalData)
library(httr)
library(dplyr)
library(stringr)
library(ggplot2)
library(MultiAssayExperiment)
library(S4Vectors)
library(UpSetR)
```


```{r}
## list all Assays/Experiments for this particular study:
Assays_available = molecularProfiles(api = cbio,
                                     studyId = 'luad_tcga',
                                     projection = 'SUMMARY')
Assay_Ids = Assays_available$molecularProfileId # in this case we have 15 Assays available
Assay_Ids

```
```{r}
##downloads and create an MAE automatically:
LUAD_Multiassay = cBioDataPack(cancer_study_id = 'luad_tcga')
#LUAD_Multiassay
```














Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
